---
title: "Data analysis"
editor: visual
execute: 
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
#| message: false
data <- read_csv("data/vazgen.csv")
main_data <- data |>
  filter(entry_id >= 1331 & entry_id <= 10021)
```

```{r}
order <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
main_data |>
  drop_na(direction, speed) |>
  mutate(direction = factor(direction, levels = order)) |>
  ggplot(aes(x = as.factor(direction), fill = direction)) + 
  geom_bar(width = 1) +
  coord_polar(start = -0.4) +
  labs(title = "Wind Rose", x = "Wind Direction", y = "Frequency", fill = "Direction") +
  theme_minimal()

```

According to this graph, the majority of winds in Yerevan blow from the south (particurarly from south-west), from Ararat through the Ararat Valley.

```{r}
main_data |> 
  mutate(mycolor = ifelse(temperature > 0, "type1", "type2")) |>
  drop_na(temperature) |>
  ggplot(aes(x = time, y = temperature)) +
  geom_segment(aes(x = time, xend = time, y = 0, yend = temperature, color = mycolor), linewidth = 0.3, alpha = 0.9) +
  theme_light() +
  theme(legend.position = "none", panel.border = element_blank()) +
  labs(title = "Changes of temperature over time",
       x = "Time", 
       y = "Temperature")
```

According to this graph, the temperature in Yerevan in winter is mostly positive, but within normal limits.
Here are average temperatures in Yerevan in winter generally for better understanding of this graph
December 5° / -2°
January 2° / -5°
February 5° / -3

```{r}
dec <- main_data |>
  filter(entry_id <= 4294)
jan <- main_data |>
  filter(entry_id >=4295, entry_id < 7270)
feb <- main_data |>
  filter(entry_id >= 7270)
dec_pm1 <- mean(dec$pm1, na.rm = TRUE)
dec_pm2_5 <- mean(dec$pm2_5, na.rm = TRUE)
dec_pm10 <- mean(dec$pm10, na.rm = TRUE)
jan_pm1 <- mean(jan$pm1, na.rm = TRUE)
jan_pm2_5 <- mean(jan$pm2_5, na.rm = TRUE)
jan_pm10 <- mean(jan$pm10, na.rm = TRUE)
feb_pm1 <- mean(feb$pm1, na.rm = TRUE)
feb_pm2_5 <- mean(feb$pm2_5, na.rm = TRUE)
feb_pm10 <- mean(feb$pm10, na.rm = TRUE)
df <- data.frame(
  Category = c("Dec", "Jan", "Feb"),
  pm1 = c(dec_pm1, jan_pm1, feb_pm1),
  pm2_5 = c(dec_pm2_5, jan_pm2_5, feb_pm2_5),
  pm3 = c(dec_pm10, jan_pm10, feb_pm10)
)
df_long <- pivot_longer(df, cols = starts_with("pm"), names_to = "pm", values_to = "Value")
order <- c("Dec", "Jan", "Feb")
df_long |>
  mutate(Category = factor(Category, levels = order)) |>
  ggplot(aes(x = Category, y = Value, fill = pm)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average air pollution in Yerevan in winter according to different particles",
       x = "Month",
       y = NULL, fill = "Pm") +
  theme_minimal()
```

In December pm1 is a little bit above of 50, which is considered as moderate, though its remaining values are good.

In all 3 months of winter values of pm2_5 are above 56, which is considered as unhealthy.

In all 3 months of winter values of pm10 are above 54, which is considered as moderate.

Below are description of pm1, pm2_5, pm10 for better understanding

PM1

These particles are so small they can bypass the body's natural defenses and penetrate deep into the lungs and bloodstream.
They can cause respiratory problems, heart disease, and even cancer.
Children and people with existing respiratory conditions are particularly vulnerable to the harmful effects of PM1.0.

PM2_5

These particles are also small enough to cause respiratory problems, but their larger size prevents them from reaching the deepest parts of the lungs.
However, they can still irritate the airways and trigger asthma attacks.
PM2.5 exposure has also been linked to an increased risk of cardiovascular disease.

PM10

These particles are larger and less likely to be inhaled deeply.
However, they can still irritate the eyes, nose, and throat.
Long-term exposure to PM10 can also lead to respiratory problems and heart disease.

```{r}
main_data |>
  drop_na(humidity) |>
  ggplot(aes(x = humidity)) +
  geom_histogram(fill = "#69b3a2", color = "#e9ecef", alpha = 0.5, binwidth = 2) +
  labs(title = "Humidity levels in winter in Yerevan",
       x = "Humidity percentage",
       y = NULL)
```

Mostly humidity % is above 60, which is normal for winter in Yerevan.